### 2023-09-19 doc/_bayes_freq.qmd
update SI to poisson-multinomial model

We begin with the probability of the frequency values given the raw barcode
reads. The first assumption we make is that the inference of the frequency
values for time $t$ is independent of any other time. Therefore, we can write
the joint probability distribution as a product of independent distributions of
the form
$$
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) =
\prod_{t=1}^T \pi(\underline{f}_t \mid \underline{r}_t).
$${#eq-freq_indep}

Writing Bayes' theorem for this series of independent terms results in
$$
\pi(\underline{f}_t \mid \underline{r}_t) \propto
\pi(\underline{r}_t \mid \underline{f}_t)
\pi(\underline{f}_t).
$${#eq-bayes_freq}
To model the prior uncertainty in the value of the frequencies, we assume a very
simple model. Notice that the frequencies at any time point must satisfy the
normalization constraint
$$
\sum_{b = 1}^B f_t^{(b)} = 1, \;\; \forall \; f_t^{(b)} \in [0, 1].
$${#eq-freq_norm}
This means that the probability distribution of the $B$ frequency values lies on
the $B$-simplex, i.e., the $B$ dimensional space constrained such that the sum
of all the dimensions must add up to one. A natural selection for a probability
that satisfies the constraint in @eq-freq_norm is the so-called Dirichlet 
distribution. The Dirichlet distribution is parametrized by a vector
$$
\underline{\alpha}_t = (\alpha^{(1)}_t, \cdots, \alpha^{(B)}_t)^\dagger, 
\;\; \alpha_t^{(k)} \geq 0 \;\; \forall \; k.
$${#eq-alpha_dir}
Given these parameters, we choose as a prior distribution for the barcode
frequency values $\pi(\underline{f}_t)$ a Dirichlet distribution of the form
$$
\pi(\underline{f}_t) = \frac{
    \Gamma\left(\sum_{b=1}^B\alpha_t^{(b)}\right)
}{
    \prod_{b=1}^B\Gamma(\alpha_t^{(b)})
} 
\prod_{b=1}^B (f_t^{(b)})^{\alpha_t^{(b)}-1}.
$${#eq-dirichlet}
@eq-dirichlet can be much more compactly written as
$$
\underline{f}_t \sim \operatorname{Dirichlet}(\underline{\alpha}_t),
$${#eq-freq_prior}
where the symbol "$\sim$" is read "distributed as". We are free to choose the
value of the parameters $\underline{\alpha}_t$ according to our prior
information. A reasonable choice is to assume that before observing any data at
all, we have no reason to believe any barcode is in higher abundance compared to
any other. This might not be entirely true at the beginning of the experiment,
where we expect the neutral barcodes to be in much higher frequency compared to
the mutant barcodes. Nevertheless, this might not be true for the following time
points, making our "uniform prior over the simplex" a reasonable choice. 
Therefore, we define
$$
\alpha_t^{(b)} = 1 \;\; \forall \; b.
$${#eq-alpha_choice}

For the likelihood term $\pi(\underline{r}_t \mid \underline{f}_t)$, we assume a
very simple multinomial model. The story behind this choice goes as follows:
Given a "true" underlying barcode frequency, when the sequencing machine reads a
particular barcode, this is equivalent to having thrown a $B$-sided die, where
the face of the die indicates the barcode that is read. This die is then rolled
$R_t = \sum_b r_t^{(b)}$ times, giving the values of the reads we observe at
time $t$. Notice that this model does not account for noise that might come from
sources such as genetic drift, DNA extraction, and PCR amplification. Although
these sources of noise are obviously important, it is not clear that we have
enough statistical power to make meaningful statements about them given our data
(more on this later on). We can then compactly write the likelihood function as
$$
\underline{r}_t \mid \underline{f}_t \sim
\operatorname{Multinomial}(\underline{f}_t).
$${#eq-freq_likelihood}

This simple Multinomial-Dirichlet model for the probability of the frequency
values is analytically tractable. Specifically, the Dirichlet distribution is
the so-called *conjugate prior* to the multinomial distribution. This means the
posterior distribution $\pi(\underline{f}_t \mid \underline{r}_t)$ has a
closed-form solution of the same form as the prior $\pi(\underline{f}_t)$, but
with updated parameters. In other words, we can write (without proof) the
posterior distribution as
$$
\underline{f}_t \mid \underline{r}_t \sim
\operatorname{Dirichlet}(\underline{\beta}_t),
$${#eq-freq_posterior}
where the new parameters are of the form
$$
\beta_t^{(b)} = \alpha_t^{(b)} + r_t^{(b)}.
$$
Simply put, when the likelihood function is multinomial, and the prior 
distribution is Dirichlet, the posterior distribution is also Dirichlet, where
the parameters of the posterior are the parameters of the prior plus the number
of observations of the corresponding output. This fact will become handy
as we build the entire posterior distribution for our inference as defined in
@eq-split_posterior.

>>>>

We begin with the probability of the frequency values given the raw barcode
reads. The first assumption is that the inference of the frequency
values for time $t$ is independent of any other time. Therefore, we can write
the joint probability distribution as a product of independent distributions of
the form
$$
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) =
\prod_{t=1}^T \pi(\underline{f}_t \mid \underline{r}_t),
$${#eq-freq_indep}
where $\underline{f}_t$ and $\underline{r}_t$ are the $t$-th row of the matrix
containing all of the measurements for time $t$. We imagine that when the 
barcode reads are obtained via sequencing, the quantified number of reads is a
Poisson sample from the "true" underlying number of barcodes within the pool. 
This translates to assuming that the number of reads for each barcode at any
time point $r^{(b)}_t$ is an independent Poisson random variable, i.e.,
$$
r^{(b)}_t \sim \text{Poiss}(\lambda^{(b)}_t),
$${#eq-freq_poisson_reads}
where the symbol "$\sim$" is read "distributed as." Furthermore, for a Poisson
distribution, we have that
$$
\lambda^{(b)}_t = \left\langle r^{(b)}_t \right\rangle = 
\left\langle 
    \left( r^{(b)}_t - \left\langle r^{(b)}_t \right\rangle \right)^2
\right\rangle,
$${#eq-freq_poisson_lambda}
where $\left\langle \cdot \right\rangle$ is the expected value. In other words
the Poisson parameter is equal to the mean and variance of the distribution. The
Poisson distribution has the convenient property that for two Poisson 
distributed random variables $X \sim \text{Poiss}(\lambda_x)$ and 
$Y \sim \text{Poiss}(\lambda_y)$, we have that
$$
Z \equiv X + Y \sim \text{Poiss}(\lambda_x + \lambda_y).
$${#eq-freq_additivity}
This additivity allows us to write the total number of reads at time $t$ $n_t$
also as a Poisson-distributed random variable of the form
$$
n_t \sim \text{Poiss}\left( \sum_{b=1}^B \lambda^{(b)}_t \right),
$${#eq-freq_poisson_total}
where the sum is taken over all $B$ barcodes.

If the total number of reads is given by @eq-freq_poisson_total, the array with
the number of reads for each barcode at time $t$, $\underline{r}_t$ is then
distributed as
$$
\underline{r}_t \sim \text{Multinomial}(n_t, \underline{f}_t),
$${#eq-freq_multinomial}
where each of the $B$ entries of the frequency vector $\underline{f}_t$ is a
function of the $\underline{\lambda}_t$ vector, given by
$$
f_t^{(b)} \equiv f_t^{(b)}(\underline{\lambda}_t) = 
\frac{\lambda_t^{(b)}}{\sum_{b'=1}^B \lambda_t^{(b')}}.
$${#eq-freq_freq_lambda}
In other words, we can think of the $B$ barcode counts as independent Poisson
samples or as a single multinomial draw with a random number of total draws,
$n_t$, and the frequency vector $\underline{f}_t$ we are interested in. Notice
that @eq-freq_freq_lambda is a deterministic function that connects the Poisson
parameters to the frequencies. Therefore, we have the equivalence that
$$
\pi(\underline{f}_t \mid \underline{r}_t) = 
\pi(\underline{\lambda}_t \mid \underline{r}_t),
$${#eq-freq_f_lambda_equiv}
meaning that the uncertainty comes from the $\underline{\lambda}_t$ vector. By
Bayes theorem, we therefore write
$$
\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
\pi(n_t, \underline{r}_t \mid \underline{\lambda}_t) \pi(\underline{\lambda}_t),
$${#eq-freq_lambda_bayes}
where we explicitly include the dependence on $n_t$. This does not affect the
distribution or brings more uncertainty because $\underline{r}_t$ already 
contains all the information to compute $n_t$ since
$$
n_t = \sum_{b=1}^B r_t^{(b)}.
$${#eq-freq_n_sum_r}
But adding the variable allows us to factorize @eq-freq_lambda_bayes as
$$
\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
\pi(n_t, \underline{r}_t \mid \underline{\lambda}_t) 
\pi(n_t \mid \underline{\lambda}_t) \pi(\underline{\lambda}_t).
$${#eq-freq_lambda_bayes_factorized}
We then have
$$
\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
\text{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)).
$${#eq-freq_r_bayes}
Furthermore, we have
$$
n_t \mid \underline{\lambda}_t \sim 
\text{Poiss}(\sum_{b=1}^B \lambda_t^{(b)}).
$${#eq=freq_n_bayes}
Finally, for our prior $\pi(\underline{\lambda}_t)$, we first assume each 
parameter is independent, i.e.,
$$
\pi(\underline{\lambda}_t) = \prod_{b=1}^B \pi(\lambda_t^{(b)}).
$$
A reasonable prior for each $\lambda_t^{(b)}$ representing the expected number
of reads for barcode $b$ should span several orders of magnitude. Furthermore,
we assume that no barcode in the dataset ever goes extinct. Thus, no frequency
can equal zero, facilitating the computation of the log frequency ratios
needed to infer the relative fitness. The log-normal distribution satisfies 
these constraints; therefore, for the prior, we assume
$$
\lambda_t^{(b)} \sim 
\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
$$
with $\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}$ as the user-defined 
parameters that characterize the prior distribution.
